---
title: "Week1"
output: pdf_document
---

## R basics

Now that we are set up with R and RStudio, we want to cover some of the very basics of R data manipulation, though we will not have comprehensive coverage of the R syntax. 

## Getting tables into R

The beginning of a "real data analysis" usually starts with getting a table of data loaded into R. The original data might be stored in an Microsoft Excel spreadsheet. There is luckily an easy way to export data from Excel such that it can be read by other software. Go to File > Save As..., and then choose Format: Comma Separated Values (csv). Excel will give a warning then that only a single sheet of data will be exported into the CSV file.

## Mammal sleep data

Let's start here with a CSV file of mammalian sleep data. This data is already bundled up as msleep [1] in the ggplot2 package, but here we want to first show how to read in the data from CSV format.

```{r}
setwd("C:/Study/eDx/Genomicsclass/edX_Courses")
tab <- read.csv("Data/msleep_ggplot2.csv")

# Two useful things to know are: what does the top of the dataframe 
# look like, and what are the dimensions of the dataframe?
head(tab)
dim(tab)
```

## Accessing Columns

We can get a column of data from a dataframe by typing the name of the dataframe, then the `$` symbol, then the name of the column with no spaces in between. First ask what the column names are using `colnames(tab)`, and then extract one of the columns. The column will be returned as a vector of numbers. Try using autocompletion on the column name using the TAB key on your keyboard. Type the name of the dataframe, then the `$` symbol, then the first few letters of the column, then hit TAB. This will autocomplete the name until the end of the name, or until the next character which would determine which column you are referring to. In RStudio, a popup window will show you the possible column names that match the first few letters you wrote.

```{r}
colnames(tab)

# How many hours of total sleep are there for the first animal in the table?
tab[1, "sleep_total"]
```

## Create and manipulate vectors with $c()$

Vectors can be combined using the `c` function. For example, we can add a final number, 1000, to the sleep totals:
```{r}
c(tab$sleep_total, 1000)

# Plot the data
plot(tab$brainwt, tab$sleep_total)

# with a logarithmic scale x-axis:
plot(tab$brainwt, tab$sleep_total, log = "x")

# The summary function gives the summary statistics of a set of values.
# What is the 3rd quartile of the total sleep of all the animals?
summary(tab$sleep_total)
```

## Subsetting/Indexing

There are a number of ways to subset a dataframe or a vector. We can use numbers to indicate the rows we want, or a logical vector which is as long as the number of rows of the dataframe, or as long as the vector. We use the square brackets [] and inside give the rows and columns we want to index, separated by a comma. If we leave out the number before or after the comma, it means, give us all the rows or all the columns.

```{r}
# Subsetting a dataframe to the first two rows:
tab[c(1, 2), ]
# The rows where the total sleep is greater than 18 hours:
tab[ tab$sleep_total > 18, ]
# The first two elements can be subset like so:
tab$sleep_total[ c(1,2) ]
# What is the average total sleep for the animals with total sleep 
# greater than 18 hours?
mean(tab$sleep_total[tab$sleep_total > 18])
```

### which()

The function which() gives us the numeric index that satisfies a logical question:

```{r}
tab$sleep_total[which(tab$sleep_total > 18)][1]

# What is the row number of the animal which has more than 18 hours of 
# total sleep and less than 3 hours of REM sleep?
which(tab$sleep_total > 18 & tab$sleep_rem < 3)

# sort() simply gives back the list of numeric values after sorting them:
sort(tab$sleep_total)

# order()
# What is the index of the animal (so the row number) with the least 
# total sleep?
order(tab$sleep_total)[1]

# rank()
rank(c(1,2,2,3))
# What's the rank of the animal in the first row of the table in terms 
# of total sleep?
rank(tab$sleep_total)

# The match() function in R is useful to find the index of the first 
# match of a vector in a second vector. We can give match() a number 
# of queries at once:
match(c("Cow","Owl monkey","Cheetah"), tab$name)

# reorder the tab dataframe to give the rows for Cow, Owl monkey and 
# Cheetah, using the vector returned by match() as a row index:
idx = match(c("Cow","Owl monkey","Cheetah"), tab$name)
tab[idx, ]

# What is the row number for "Cotton rat" in the tab dataframe?
match("Cotton rat", tab$name)

# Factors
# The levels of a factor and their order can be seen using the levels()
# function. Let's see an example:
vec = c("red","blue","red","green","green","yellow","orange")
fac <- factor(vec)
fac
levels(fac)

# reorder the levels
fac2 = factor(vec, levels=c("blue","green","yellow","orange","red"))
fac2
levels(fac2)

# table()
# How many rodents (Rodentia) are in the table?
table(tab$order)

# split() is a function which takes a vector and splits it into a list, 
# by grouping the vector according to a factor. 
s <- split(tab$sleep_total, tab$order)
# What is the mean hours of total sleep of the rodents?
mean(s[["Rodentia"]])

# lapply() and sapply() are useful functions for applying a function
# repeatedly to a vector or list. lapply() returns a list, while sapply()
# tries to "simplify", returning a vector if possible (if there is only 
# one element returned by the function for each element of the input.
lapply(s, mean)
sapply(s, mean)
tapply(tab$sleep_total, tab$order, mean)

# What is the standard deviation of total hours of sleep for the 
# Primates Order?
tapply(tab$sleep_total, tab$order, sd)
